---
title: "Exercises 3 - apply Solutions"
output:
  github_document
---

We will use a dataset from the UCI Machine Learning Repository.  This data set is about wine quality (we'll consider the red wine data).  You can learn more about the data [here](https://archive.ics.uci.edu/ml/datasets/Wine+Quality).  

The data description describes the following variables:  

> Input variables (based on physicochemical tests)  
1 - fixed acidity  
2 - volatile acidity  
3 - citric acid  
4 - residual sugar  
5 - chlorides  
6 - free sulfur dioxide  
7 - total sulfur dioxide  
8 - density  
9 - pH  
10 - sulphates  
11 - alcohol  

>Output variable (based on sensory data):  
12 - quality (score between 0 and 10)  

Read in the data using the code below.

```{r}
wineData <- readr::read_delim("winequality-red.csv", delim = ";")
```

1. Find the column means.

2. Find the column standard deviations.  

3. Use the `apply()` function along with the `summary()` function to find summaries about the columns.

4. 
```{r}

```


<hr>

<br>

<a href = "https://jbpost2.github.io/Improving-R-Programs/CourseFiles.html"><button type="button">Course Files Page</button></a>
