---
title: "Exercises 3 - apply"
output:
  github_document
---

We will use a dataset from the UCI Machine Learning Repository.  This data set is about wine quality (we'll consider the red wine data).  You can learn more about the data [here](https://archive.ics.uci.edu/ml/datasets/Wine+Quality).  

The data description describes the following variables:  

> Input variables (based on physicochemical tests)  
1 - fixed acidity  
2 - volatile acidity  
3 - citric acid  
4 - residual sugar  
5 - chlorides  
6 - free sulfur dioxide  
7 - total sulfur dioxide  
8 - density  
9 - pH  
10 - sulphates  
11 - alcohol  

>Output variable (based on sensory data):  
12 - quality (score between 0 and 10)  

Read in the data using the code below.

```{r, message = FALSE, warning = FALSE}
wineData <- readr::read_delim("https://raw.githubusercontent.com/jbpost2/Improving-R-Programs/master/datasets/winequality-red.csv", delim = ";")
```

1. Find the column means.

2. Find the column standard deviations.  Note that you will need to coerce the `tibble` to a `matrix` (use `as.matrix()`) in order to use the `colSds` function from the `matrixStats` package.  (Alternatively, you can use the `apply()` function.)

3. Use the `apply()` function along with the `summary()` function to find summaries about the columns.

4. Run the code below to fit a simple linear regression model for quality using each of the variables (columns in the data set).  We'll learn more about writing functions in a bit!  

```{r}
slrFits <- apply(wineData[,-ncol(wineData)], MARGIN = 2, function(x){lm(wineData$quality ~ x)})
```

`slrFits` is a list object where each element has the results of a call to `lm()` (which itself is a list of length 12).

```{r}
str(slrFits, max.level = 1)
```

Use `lapply()` to run the `summary()` function on each SLR fit.  This should output p-values and info about each model fit.  

<hr>

<br>

<a href = "https://jbpost2.github.io/Improving-R-Programs/CourseFiles.html"><button type="button">Course Files Page</button></a>
