---
title: "Exercises 3 - apply Solutions"
output:
  github_document
---

#2. Report the means of each numeric column.
colMeans(select(myCO2, conc:group), na.rm = TRUE)

#3. Report the standard deviations of each column (use the matrixStats package).
library(matrixStats)
colSds(as.matrix(select(myCO2, conc:group)))
# alternatively:	apply(select(myCO2, conc:group), MARGIN = 2, FUN=sd)



1) Use `apply` to obtain the 5-number summary of each column of the built-in mtcars dataset.

```{r}
apply(mtcars, MARGIN = 2, summary)
```

2) Use `apply` to compute the row sums of our matrix 'mat' and print it to the console. 

```{r}
apply(mat, 1, sum)
```


3) Start by fitting a linear regression on some of the mtcars variables. What type of object is this?

```{r, echo=T, eval=T}
fit <- lm(mpg ~ wt, data=mtcars)
```

> Print the class of fit to the console. Then use `lapply` to print the class of each element of fit.

```{r}
class(fit)
lapply(fit, class)
```

> Print the length of fit to the console. Then use `sapply` to print the length of each element of fit to the console.

```{r}
length(fit)
sapply(fit, length)
```


<hr>

<br>

<a href = "https://jbpost2.github.io/Improving-R-Programs/CourseFiles.html"><button type="button">Course Files Page</button></a>
